Experiences Gamifying Developer Adoption\\ of Practices and Tools

### 2014-03-07 paper_id51.tex
addressed review comments
updated most of the comments.

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

Maehr proposes an affirmative theory whereby individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that are significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

@@

Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.

@@

Part of the research objective was to determine whether we can drive adoption of a practice through gamification.  As previously discussed, designed the game in Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

@@

In the third stage, developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The ``Info'' button provides a level indicator via colored contents and activates a chart display windows when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.  This final configuration remained as the display until the end of the pilot.

@@

We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio?

@@

To determine the correlation with edit duration, we used R\cite{Rcitation} to construct a linear regression model for unstructured navigation .  The linear model found a positive correlation with unstructured navigation with a p-value <.001 on 1694 degrees of freedom.  The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the unstructured navigation category.

@@

To evaluate whether developers improved their practices for structured navigation we established a metric, Navigation ratio, as the number as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.  Our hypothesis is that instant feedback from points combined with game information would result in an increase in the navigation ratio for developers in the pilot.

@@

In these results, we see an effect from the points display in Blaze in week three when the developers could see their points feedback for the first time.  We conducted a Wilcoxon Rank sum test on points similar to the way we did for navigation ratio to determine significance.  Across the pilot developers, the difference is not significant at the .05 level. The p-value for week two against week three is .19, and for week one against week three it is .31.  We do know from feedback discussions and usage data that two developers took notice of the points and used Sando more in the third week.  Their use of Sando continued at a lower level in weeks four and five.   The effect of points feedback is greater than navigation ratio during week three because of the additional emphasis of the ten point bonus given for Sando usage.  Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them "a lot" also supports our conclusion that two developers were influenced by the game elements in Blaze.

@@

Following our data analysis, we conducted one-on-one feedback sessions with the pilot participants. Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management. The participants received the feedback in the form of one-on one presentations of quantitative charts and written analysis of the information.  The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.

>>>>

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may not be motivated to start using them unless they interact with a peer developer who promotes them \cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

Maehr proposes an affirmative theory whereby individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that are significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  %Maehr also states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

@@

Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM (configuration management) submissions.

@@

Part of the research objective was to determine whether we can drive adoption of a practice through gamification.  As previously discussed, we designed the game in Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

@@

In the third stage, developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The ``Info'' button provides a level indicator via colored contents and activates a chart display window when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.  This final configuration remained as the display until the end of the pilot.

@@

We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio.

@@

To determine the correlation with edit duration, we used R\cite{Rcitation} to construct a linear regression model for unstructured navigation.  The linear model found a positive correlation with unstructured navigation with a p-value <.001 on 1694 degrees of freedom.  The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the unstructured navigation category.

@@

To evaluate whether developers improved their practices for structured navigation we established a metric, Navigation ratio, as the number of structured navigation events in a session or period of time over the number of unstructured navigation events.  Our hypothesis is that instant feedback from points combined with game information would result in an increase in the navigation ratio for developers in the pilot.

@@

In these results, we see an effect from the points display in Blaze in week three when the developers could see their points feedback for the first time.  We conducted a Wilcoxon Rank sum test on points similar to the way we did for navigation ratio to determine significance.  Across the pilot developers, the difference is not significant at the .05 level. The p-value for week two against week three is .19, and for week one against week three it is .31.  We do know from feedback discussions and usage data that two developers took notice of the points and used Sando more in the third week.  Their use of Sando continued at a lower level in weeks four and five.   The effect of points feedback is greater than navigation ratio during week three because of the additional emphasis of the ten point bonus given for Sando usage.  Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them ``a lot'' also supports our conclusion that two developers were influenced by the game elements in Blaze.

@@

Following our data analysis, we conducted one-on-one feedback sessions with the pilot participants. Before initiating the discussion, we assured the individual developer that the data shared with them is confidential and we will not share the same with management. The participants received the feedback in the form of one-on-one presentations of quantitative charts and written analysis of the information.  The key quantitative chart shown in Figure \ref{fig:developercomparison} compared their distribution of time in each category to the median for the group.
